Estimating the incidence of SARS-CoV-2 from prevalence data

### 2023-06-10 cis-imperfect-testing.tex
Add literature

Comparing the simulated and CIS data suggested that this is due to the high frequency of single positive episodes in the CIS data.
A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed that in a related setting that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
\todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}

Discussion and interrogation of the data led me to believe the most likely reason for an unexpectedly high number of single positive episodes was the presence of false negatives, meaning a false negative as a negative test result returned by a detectable individual.
This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process (see \cref{E-intro:sec:PCR}).
\todo{Ref to section in intro explaining PCR testing}
Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
Consideration of these reasons are beyond the scope of this thesis, and I group all of these possible reasons together.

There are two strands of evidence that support this hypothesis.
First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{E-ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})

\emph{Intermittent negatives} are the clearest example of a false negative.
An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
I explicitly included these in the viral load model of \cref{E-ATACCC} but ignored in \cref{E-perf-test}.
Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.

@@

Each test that was previously positive, is now positive with probability $p_\text{sens}$.

@@

>>>>

Comparing the simulated and CIS data suggested that this is due to the high frequency of single positive episodes in the CIS data.
A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed, in a related setting, that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
\todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}

Discussion and interrogation of the data led me to believe the most likely reason for an unexpectedly high number of single positive episodes was the presence of false negatives, meaning a false negative as a negative test result returned by a detectable individual.
This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process (see \cref{E-intro:sec:PCR}).
Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
Consideration of these reasons are beyond the scope of this thesis, and I group all of these possible reasons together.

There are two strands of evidence that support this hypothesis.
First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{E-ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate}).

\todo[inline]{These next two paragraphs are kinda disjointed but not sure how to bring them in}

\emph{Intermittent negatives} are the clearest example of a false negative.
An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.

@@

@@

\ifSubfilesClassLoaded{
  \listoftodos
}{}

### 2023-06-10 cis-perfect-testing.tex
Add literature

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}); \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\todo{double interval censoring refs}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
A more familiar setting for this work

@@

\section{Modelling the duration}\label{perf-test:sec:model}
In this section, I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).

I model only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

>>>>

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored; approaches are reviewd by \textcite{sunAnalysis,bogaertsSurvival}.
\todo{double interval censoring refs}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

Few studies consider the combination of both double censoring (of any type) and truncation, especially within the human biostatistical literature, although the general frameworks developed in the 1970s do include this case \autocites{turnbullEmpirical}{dempsterMaximum}.
Perhaps the main use of human studies involving double interval censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
\Textcite{sunEmpirical,bacchettiNonparametric} discuss this, however, they use a conditional likelihood approach which conditions on both the truncation (a selection effect) and the time of the infection.
The reason for the latter is unclear, perhaps though the infection time is considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}; a similar approach is taken by \textcite{shenNonparametric}.
Within CIS, detected infections are more likely to occur in the period where there is frequent testing, and hence conditioning on the infection time ovestates the probability of the infection being detected.
Furthermore, the interpretation and justification of a conditional likelihood approach is unclear in a Bayesian framework.

A more common setting for this work is estimating the survival time of bird nests~\autocite{heiseyABCs}.
In this setting, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring; this is the approach I adopt in \cref{perf-test:sec:model}.
In bird nesting studies, the visit times are normally global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.

I use a Bayesian framework for this analysis for two reasons.
First, asymptotic results for the variance of frequentist estimators do not apply.
\todo{Check why asymptotic results don't work}
This complicates the construction of confidence intervals, which may require bootstrapping or other computationally intensive methods.
Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
These assumptions may be needed to combat the lack of information in the data due to the trunction~\autocite{caoBias}.

Bayesian methods have previously been applied to nest survival~\autocites{heBayesiana}{heBayesian}{caoModeling} but these augment the data with unobserved times of infection; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).
I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.

@@

\section{Modelling the duration}\label{perf-test:sec:model}
In this section, I develop a model accounting for double interval censoring and truncation.
I model only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

### 2023-06-10 thesis.tex
Add literature

\newcommand{\citePersonalComms}{(personal communication)}

@@

The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\cref{cisProtocol}.
In brief, households are invited to the study from databases held by the ONS.
The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
Within a household, all individuals aged 2 and over were invited to participate.
Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.

@@

>>>>

\newcommand\citePersonalComms[1]{(#1, personal communication)}

@@

The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
In brief, households are invited to the study from databases held by the ONS.
The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
Within a household, all individuals aged 2 and over were invited to participate.
Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.

@@

\listoftodos

\printbibliography
